---
title: "Model Comparison for Syndromic Surveillance"
subtitle: "Systematic Evaluation of Detection Algorithms: `r params$area`"
author: "Andrea Pedot"
date: "`r Sys.Date()`"
params:
  area: "Lombardia" # Options: Lombardia, Milano, Valtellina
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(mgcv)
library(surveillance)
library(kableExtra)
library(lubridate)
library(targets)
library(patchwork)
```

# Introduction

## Syndromic Surveillance Overview

Syndromic surveillance is a public health strategy that monitors pre-diagnostic health indicators to detect disease outbreaks earlier than traditional laboratory-confirmed reporting. Rather than waiting for confirmed diagnoses, syndromic surveillance systems analyze patterns in Emergency Department (ED) visits, categorized by presenting symptoms or chief complaints.
This approach enables:

- **Early detection**: Identifying unusual health patterns before laboratory confirmation
- **Situational awareness**: Real-time monitoring of population health trends
- **Resource planning**: Anticipating healthcare system demands

## Study Context

This analysis evaluates syndromic surveillance data from **`r params$area`**, covering ED visits from December 2022 through March 2026. The data encompasses nine syndrome categories representing distinct epidemiological patterns:

**High-volume, seasonal syndromes** (modeled with flexible approaches):

- `tot`: Total ED visits (baseline denominator)
- `resp`: Respiratory conditions
- `polm`: Pulmonary conditions
- `ili`: Influenza-like illness
- `gastro`: Gastrointestinal conditions

**Low-volume, sporadic syndromes** (requiring robust, conservative thresholds):
  
- `alt_cosc`: Altered consciousness
- `conv`: Convulsions
- `dec_gravi`: Unexpected deaths
- `int_farm`: Drug intoxications
- `rash_cut`: Cutaneous rashes

## Objective

This report systematically compares four statistical approaches for anomaly detection in syndromic surveillance, evaluating their performance across multiple dimensions including accuracy, calibration, sensitivity to outbreak detection, and operational characteristics.

```{r load-data}
# Load all required data from targets
clean_data <- tar_read(clean_data)
all_data_list <- tar_read(all_data_list)
all_stats_list <- tar_read(all_stats_list)
all_baselines_list <- tar_read(all_baselines_list)

# Model results
val_results <- tar_read(val_results)
farrington_results <- tar_read(farrington_results)
glrnb_results <- tar_read(glrnb_results)
poisson_results <- tar_read(poisson_results)

# Sensitivity results (now unified with multiple durations)
sens_results_gam <- tar_read(sens_results_gam)
sens_results_farrington <- tar_read(sens_results_farrington)

# Filter for current area
df_long <- all_data_list[[params$area]]

cat("Analysis Area:", params$area, "\n")
cat("Date Range:", as.character(min(df_long$Data_inizio_sett)), "to", 
    as.character(max(df_long$Data_inizio_sett)), "\n")
cat("Total Observations:", nrow(df_long))
```


# Methods

## Data Infrastructure

### Temporal Structure

The input data consists of rolling 7-day windows with daily overlap. Each row represents a week-long aggregation of ED visits, advancing by one day. This structure provides smooth temporal resolution while maintaining sufficient statistical power for rare syndromes.

```{r data-structure}
# Show data structure
df_long |>
  group_by(sindrome) |>
  summarise(
    First_Date = min(Data_inizio_sett),
    Last_Date = max(Data_inizio_sett),
    N_Observations = n(),
    Mean_Cases = round(mean(casi, na.rm = TRUE), 1),
    SD_Cases = round(sd(casi, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  kable(caption = "Data Summary by Syndrome") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Train-Test Split Strategy

To prevent data leakage and ensure valid out-of-sample evaluation, the data is partitioned as follows:

- **Training Period**: December 2022 – March 2025 (~27 months)
- **Testing Period**: March 2025 – March 2026 (12 months)

All model parameters, including seasonal patterns and adaptive calibration factors, are estimated exclusively from training data. The testing period serves as a held-out validation set for performance comparison.

```{r split-definition}
validation_start <- max(df_long$Data_inizio_sett) - days(365)

train_df <- df_long |> filter(Data_inizio_sett <= validation_start)
test_df <- df_long |> filter(Data_inizio_sett > validation_start)

cat("Training observations:", nrow(train_df), "\n")
cat("Testing observations:", nrow(test_df), "\n")
cat("Validation start date:", as.character(validation_start))
```


## Exploratory Data Analysis

### Long-Term Trends

Visualizing the temporal evolution of each syndrome reveals underlying patterns that detection algorithms must account for.

```{r eda-longterm}
ggplot(df_long, aes(x = Data_inizio_sett, y = casi, color = sindrome)) +
  geom_line(alpha = 0.6) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE, linewidth = 1) +
  facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Long-Term Syndrome Trends",
    subtitle = "Smoothed GAM trend overlaid on raw weekly counts",
    x = "Date", y = "Weekly Cases"
  )
```

### Seasonal Patterns

Annual seasonality is a critical feature for respiratory and influenza-like syndromes, with predictable winter peaks.

```{r eda-seasonal}
df_long |>
  mutate(month = month(Data_inizio_sett, label = TRUE)) |>
  group_by(sindrome, month) |>
  summarise(mean_cases = mean(casi, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month, y = mean_cases, group = sindrome, color = sindrome)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Average Seasonal Patterns by Month",
    x = "Month", y = "Mean Weekly Cases"
  )
```

### Overdispersion Analysis

Overdispersion occurs when variance exceeds the mean, violating Poisson assumptions. The overdispersion index (variance/mean) quantifies this deviation.

```{r eda-overdispersion}
stats_current <- all_stats_list[[params$area]]
overdisp_stats <- stats_current$overdispersion

overdisp_stats |>
  mutate(
    Overdispersed = ifelse(overdisp_index > 1, "Yes", "No")
  ) |>
  kable(digits = 2, caption = "Overdispersion Diagnostics by Syndrome") |>
  kable_styling(bootstrap_options = c("striped", "hover")) |>
  column_spec(5, bold = TRUE, 
              color = ifelse(overdisp_stats$overdisp_index > 1, "red", "green"))
```

### Autocorrelation Structure

Temporal autocorrelation indicates the degree to which consecutive observations are correlated, which affects both model specification and threshold calibration.

```{r eda-acf}
acf_data <- stats_current$acf

ggplot(acf_data, aes(x = lags, y = acf_vals, fill = sindrome)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "red", alpha = 0.5) +
  facet_wrap(~sindrome, ncol = 3, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Autocorrelation Function (ACF) of Weekly Changes",
    subtitle = "Dashed lines indicate approximate 95% significance bounds",
    x = "Lag (weeks)", y = "ACF"
  )
```


## Model Specifications

### Model 1: Adaptive GAM with Negative Binomial Distribution

A Generalized Additive Model using thin-plate regression splines for trend and cyclic splines for seasonality, with a 14-day adaptive bias correction.

**Flexible specification** (seasonal syndromes):
$$\log(\mu_t) = \beta_0 + f_1(\text{week\_num}) + f_2(\text{week\_of\_year}) + \gamma \cdot \text{dow}$$

**Stiff specification** (sporadic syndromes):
$$\log(\mu_t) = \beta_0 + f_1(\text{week\_num}) + \gamma \cdot \text{dow}$$

where $f_1$ uses thin-plate splines with $k=20$ or $k=3$ knots, and $f_2$ uses cyclic cubic splines with $k=12$.

The 14-day adaptive calibration applies a multiplicative bias factor:
$$\text{bias} = \frac{\bar{y}_{t-14:t}}{\bar{\hat{y}}_{t-14:t}}$$

### Model 2: Farrington Flexible Algorithm

The Noufaily-adjusted Farrington algorithm, implemented in the `surveillance` R package, uses quasi-Poisson regression with historical reference windows.

Key parameters:

- Reference window: $b = 2$ years, $w = 21$ days (3 weeks)
- Reweighting for past outbreaks: `weightsThreshold = 2.58`
- Trend inclusion: `trend = TRUE`
- Power transformation: $y^{2/3}$

### Model 3: GLR for Negative Binomial (CUSUM-based)

The Generalized Likelihood Ratio statistic for detecting shifts in the mean of a Negative Binomial distribution, using a CUSUM-type accumulation.

$$\Lambda_t = \max\left(0, \Lambda_{t-1} + \log\frac{f(y_t; \mu_1, \alpha)}{f(y_t; \mu_0, \alpha)}\right)$$

An alarm is triggered when $\Lambda_t > h$ (threshold controlling ARL).

### Model 4: Poisson Baseline (Benchmark)

A naive benchmark using the historical training mean as the expected value:
$$\hat{\mu} = \frac{1}{n_{\text{train}}} \sum_{t \in \text{train}} y_t$$

Thresholds are set using Poisson quantiles. This model serves as a lower-bound benchmark.


# Results

## Model Training and Residual Diagnostics

### GAM Residual Analysis

```{r gam-standardize}
# Standardize GAM results
gam_std <- standardize_gam_results(val_results, params$area)
```

#### Q-Q Plot (GAM)

```{r gam-qq}
gam_std |>
  group_by(sindrome) |>
  mutate(
    observed = sort(Std_Residual),
    theoretical = qnorm(ppoints(n()))
  ) |>
  ungroup() |>
  ggplot(aes(x = theoretical, y = observed)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick") +
  facet_wrap(~sindrome, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(
    title = "Q-Q Plot: GAM Standardized Residuals",
    subtitle = "Deviation from diagonal indicates departure from normality",
    x = "Theoretical Quantiles", y = "Sample Quantiles"
  )
```

#### Calibration Plot (GAM)

```{r gam-calibration}
ggplot(gam_std, aes(x = Expected, y = Observed)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "gam", color = "firebrick", se = FALSE) +
  facet_wrap(~sindrome, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(
    title = "Calibration: GAM Observed vs. Expected",
    subtitle = "Perfect calibration follows the diagonal; red line shows actual trend",
    x = "Expected Cases", y = "Observed Cases"
  )
```

#### Residual Autocorrelation (GAM)

```{r gam-residual-acf}
gam_diag <- calculate_residual_diagnostics(gam_std, "Residual")

ggplot(gam_diag$acf, aes(x = lags, y = acf_vals)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "red", alpha = 0.5) +
  facet_wrap(~sindrome, ncol = 3) +
  theme_minimal() +
  labs(
    title = "Residual Autocorrelation: GAM Model",
    subtitle = "Residual ACF should be near zero if model captures temporal structure",
    x = "Lag", y = "ACF"
  )
```

### Farrington Residual Analysis

```{r farrington-standardize}
farr_std <- standardize_farrington_results(farrington_results, params$area)
```

```{r farrington-plots}
#| layout-ncol: 1

if (nrow(farr_std) > 0 && sum(!is.na(farr_std$Expected)) > 10) {
  
  # Q-Q Plot
  p1 <- farr_std |>
    filter(!is.na(Std_Residual)) |>
    group_by(sindrome) |>
    mutate(
      observed = sort(Std_Residual),
      theoretical = qnorm(ppoints(n()))
    ) |>
    ungroup() |>
    ggplot(aes(x = theoretical, y = observed)) +
    geom_point(alpha = 0.3, color = "darkorange") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick") +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal() +
    labs(title = "Q-Q Plot: Farrington Standardized Residuals",
         x = "Theoretical Quantiles", y = "Sample Quantiles")
  
  print(p1)
  
  # Calibration Plot
  p2 <- farr_std |>
    filter(!is.na(Expected)) |>
    ggplot(aes(x = Expected, y = Observed)) +
    geom_point(alpha = 0.2, color = "darkorange") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_smooth(method = "gam", color = "firebrick", se = FALSE) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal() +
    labs(title = "Calibration: Farrington Observed vs. Expected",
         x = "Expected Cases", y = "Observed Cases")
  
  print(p2)
  
} else {
  cat("Insufficient Farrington results for diagnostic plots.")
}
```

### Poisson Baseline Residual Analysis

```{r poisson-plots}
poisson_df <- poisson_results[[params$area]]

if (!is.null(poisson_df) && nrow(poisson_df) > 0) {
  
  # Q-Q Plot
  poisson_df |>
    group_by(sindrome) |>
    mutate(
      observed = sort(Residual),
      theoretical = qnorm(ppoints(n()))
    ) |>
    ungroup() |>
    ggplot(aes(x = theoretical, y = observed)) +
    geom_point(alpha = 0.3, color = "forestgreen") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick") +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal() +
    labs(
      title = "Q-Q Plot: Poisson Baseline Residuals",
      x = "Theoretical Quantiles", y = "Sample Quantiles"
    )
}
```


## Performance Metrics Comparison

```{r metrics-calculation}
# Calculate metrics for each model
metrics_gam <- calculate_model_metrics(gam_std, "GAM Adaptive")

metrics_farr <- if (nrow(farr_std) > 0) {
  calculate_model_metrics(farr_std, "Farrington")
} else {
  tibble()
}

metrics_poisson <- if (!is.null(poisson_df) && nrow(poisson_df) > 0) {
  calculate_model_metrics(poisson_df, "Poisson Baseline")
} else {
  tibble()
}

# GLR-NB metrics (simplified - uses alarm-based evaluation)
glrnb_df <- glrnb_results[[params$area]]
metrics_glrnb <- if (!is.null(glrnb_df) && nrow(glrnb_df) > 0) {
  glrnb_df |>
    group_by(sindrome) |>
    summarise(
      Model = "GLR-NB",
      N_Obs = n(),
      RMSE = sqrt(mean((Observed - Expected)^2, na.rm = TRUE)),
      MAE = mean(abs(Observed - Expected), na.rm = TRUE),
      MAPE = mean(abs(Observed - Expected) / (Observed + 1), na.rm = TRUE) * 100,
      Mean_Bias = mean(Observed - Expected, na.rm = TRUE),
      R2 = NA_real_,
      Coverage_95 = NA_real_,
      Coverage_99 = NA_real_,
      Alert_Rate_95 = NA_real_,
      Alert_Rate_99 = mean(Alarm == 1, na.rm = TRUE) * 100,
      .groups = "drop"
    )
} else {
  tibble()
}

# Combine all metrics
all_metrics <- bind_rows(metrics_gam, metrics_farr, metrics_poisson, metrics_glrnb) |>
  filter(!is.na(RMSE))
```

### Summary Statistics

```{r metrics-summary}
all_metrics |>
  group_by(Model) |>
  summarise(
    Avg_RMSE = mean(RMSE, na.rm = TRUE),
    Avg_MAE = mean(MAE, na.rm = TRUE),
    Avg_R2 = mean(R2, na.rm = TRUE),
    Avg_Coverage_95 = mean(Coverage_95, na.rm = TRUE),
    Avg_Alert_Rate_99 = mean(Alert_Rate_99, na.rm = TRUE),
    .groups = "drop"
  ) |>
  kable(digits = 2, caption = "Model Performance Summary (Averaged Across Syndromes)") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Detailed Metrics by Syndrome

```{r metrics-detailed}
all_metrics |>
  select(sindrome, Model, RMSE, MAE, R2, Coverage_95, Alert_Rate_99) |>
  pivot_wider(
    names_from = Model,
    values_from = c(RMSE, MAE, R2, Coverage_95, Alert_Rate_99),
    names_glue = "{Model}_{.value}"
  ) |>
  kable(digits = 2, caption = "Detailed Performance Metrics by Syndrome") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 10) |>
  scroll_box(width = "100%")
```

### RMSE Comparison Visualization

```{r metrics-plot}
all_metrics |>
  ggplot(aes(x = reorder(sindrome, RMSE), y = RMSE, fill = Model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(
    title = "RMSE Comparison Across Models",
    x = "Syndrome", y = "Root Mean Square Error"
  )
```


## Historical Alert Reconstruction

### GAM Alerts

```{r gam-alerts}
gam_alerts <- gam_std |>
  filter(Status == "RED") |>
  select(Date, Syndrome = sindrome, Observed, Expected, Status) |>
  mutate(Model = "GAM Adaptive") |>
  arrange(Syndrome, desc(Date))

if (nrow(gam_alerts) > 0) {
  gam_alerts |>
    head(100) |>
    kable(digits = 1, caption = "GAM Adaptive: Critical Alerts (99% Threshold)") |>
    kable_styling(bootstrap_options = c("striped", "condensed", "hover")) |>
    scroll_box(height = "400px")
} else {
  cat("No critical alerts detected by GAM model.")
}
```

### Farrington Alerts

```{r farrington-alerts}
farr_alerts <- farr_std |>
  filter(grepl("RED", Status)) |>
  select(Date, Syndrome = sindrome, Observed, Expected, Status) |>
  mutate(Model = "Farrington") |>
  arrange(Syndrome, desc(Date))

if (nrow(farr_alerts) > 0) {
  farr_alerts |>
    head(100) |>
    kable(digits = 1, caption = "Farrington: Critical Alerts (99% Threshold)") |>
    kable_styling(bootstrap_options = c("striped", "condensed", "hover")) |>
    scroll_box(height = "400px")
} else {
  cat("No critical alerts detected by Farrington model.")
}
```

### GLR-NB Alerts

```{r glrnb-alerts}
if (!is.null(glrnb_df) && nrow(glrnb_df) > 0) {
  glrnb_alerts <- glrnb_df |>
    filter(Alarm == 1) |>
    select(Date, Syndrome = sindrome, Observed, Expected, Status) |>
    mutate(Model = "GLR-NB") |>
    arrange(Syndrome, desc(Date))
  
  if (nrow(glrnb_alerts) > 0) {
    glrnb_alerts |>
      head(100) |>
      kable(digits = 1, caption = "GLR-NB: Detected Alarms") |>
      kable_styling(bootstrap_options = c("striped", "condensed", "hover")) |>
      scroll_box(height = "400px")
  } else {
    cat("No alarms detected by GLR-NB model.")
  }
}
```

### Alert Concordance

```{r alert-concordance}
# Combine all alerts
all_alerts <- bind_rows(
  gam_alerts |> select(Date, Syndrome, Model),
  farr_alerts |> select(Date, Syndrome, Model),
  if (exists("glrnb_alerts") && nrow(glrnb_alerts) > 0) {
    glrnb_alerts |> select(Date, Syndrome, Model)
  } else {
    tibble()
  }
)

if (nrow(all_alerts) > 0) {
  # Count alerts by date and syndrome
  concordance <- all_alerts |>
    group_by(Date, Syndrome) |>
    summarise(
      Models_Alerting = paste(sort(unique(Model)), collapse = ", "),
      N_Models = n_distinct(Model),
      .groups = "drop"
    ) |>
    filter(N_Models >= 2) |>
    arrange(desc(N_Models), desc(Date))
  
  if (nrow(concordance) > 0) {
    concordance |>
      head(50) |>
      kable(caption = "Alert Concordance: Dates with Multiple Models Alerting") |>
      kable_styling(bootstrap_options = c("striped", "hover"))
  } else {
    cat("No dates where multiple models simultaneously triggered alerts.")
  }
}
```


## Sensitivity Analysis: Probability of Detection

### GAM Model - Multi-Duration POD

```{r sens-gam}
sens_gam_df <- sens_results_gam[[params$area]]

if (!is.null(sens_gam_df) && nrow(sens_gam_df) > 0) {
  ggplot(sens_gam_df, aes(x = Total_Cases, y = POD, color = factor(Duration_Days))) +
    geom_hline(yintercept = 0.8, linetype = "dotted", color = "gray40") +
    geom_line(linewidth = 1) +
    geom_point(size = 1.5, alpha = 0.7) +
    scale_x_log10(labels = scales::comma) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
    scale_color_brewer(palette = "Dark2", name = "Outbreak\nDuration (days)") +
    facet_wrap(~sindrome, ncol = 3) +
    theme_minimal() +
    labs(
      title = "GAM Adaptive: Sensitivity Analysis",
      subtitle = "Probability of Detection vs. Outbreak Magnitude (4, 14, 49-day durations)",
      x = "Total Outbreak Cases (log scale)", y = "Probability of Detection"
    )
} else {
  cat("GAM sensitivity results not available.")
}
```

### Farrington Model - Multi-Duration POD

```{r sens-farrington}
sens_farr_df <- sens_results_farrington[[params$area]]

if (!is.null(sens_farr_df) && nrow(sens_farr_df) > 0) {
  ggplot(sens_farr_df, aes(x = Total_Cases, y = POD, color = factor(Duration_Days))) +
    geom_hline(yintercept = 0.8, linetype = "dotted", color = "gray40") +
    geom_line(linewidth = 1) +
    geom_point(size = 1.5, alpha = 0.7) +
    scale_x_log10(labels = scales::comma) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
    scale_color_brewer(palette = "Dark2", name = "Outbreak\nDuration (days)") +
    facet_wrap(~sindrome, ncol = 3) +
    theme_minimal() +
    labs(
      title = "Farrington Flexible: Sensitivity Analysis",
      subtitle = "Probability of Detection vs. Outbreak Magnitude (4, 14, 49-day durations)",
      x = "Total Outbreak Cases (log scale)", y = "Probability of Detection"
    )
} else {
  cat("Farrington sensitivity results not available.")
}
```

### Comparative POD at 80% Detection Threshold

```{r sens-comparison}
# Calculate the minimum outbreak size for 80% detection
get_pod80_threshold <- function(df, model_name) {
  if (is.null(df) || nrow(df) == 0) return(tibble())
  
  df |>
    group_by(sindrome, Duration_Days) |>
    arrange(Total_Cases) |>
    filter(POD >= 0.8) |>
    slice_head(n = 1) |>
    ungroup() |>
    mutate(Model = model_name) |>
    select(sindrome, Duration_Days, Threshold_80_POD = Total_Cases, Model)
}

pod80_gam <- get_pod80_threshold(sens_gam_df, "GAM Adaptive")
pod80_farr <- get_pod80_threshold(sens_farr_df, "Farrington")

pod80_combined <- bind_rows(pod80_gam, pod80_farr)

if (nrow(pod80_combined) > 0) {
  pod80_combined |>
    pivot_wider(names_from = Model, values_from = Threshold_80_POD) |>
    kable(caption = "Minimum Outbreak Size for 80% Detection Probability") |>
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```


# Conclusion

## Data Leakage Prevention

This analysis implements strict temporal separation to prevent information leakage:

1. **Training/Validation Split**: All model parameters are estimated using data up to March 2025. The subsequent 12 months serve exclusively as the test set.

2. **Rolling-Origin Cross-Validation**: For validation metrics, models are re-trained at each time point using only past data, simulating real-world deployment conditions.

3. **Adaptive Calibration Isolation**: The 14-day bias correction uses only observations prior to each prediction point, never incorporating future information.

## Uncertainty Quantification

Uncertainty is addressed through multiple mechanisms:

- **Negative Binomial Distribution**: Explicitly models overdispersion, providing wider prediction intervals than Poisson
- **Monte Carlo Simulation**: Sensitivity analysis uses n=50 random trials per magnitude level to account for temporal variability
- **Dual Thresholds**: 95% (Yellow) and 99% (Red) confidence levels provide graduated alert severity

## Generalizability Considerations

**Strengths**:

- Models are area-specific, capturing local patterns
- Syndrome-specific model selection (flexible vs. stiff) adapts to epidemiological characteristics
- Three geographic areas (Lombardia, Milano, Valtellina) provide internal replication

**Limitations**:

- Results may not generalize to regions with different healthcare infrastructure or population demographics
- Rare syndromes have limited statistical power for POD estimation
- The rolling-window data structure introduces autocorrelation that may inflate apparent performance

## General Limitations

1. **No Ground Truth**: Without confirmed outbreak labels, we cannot directly measure true positive/false positive rates

2. **Computational Constraints**: Sensitivity analysis is limited to 50 trials per condition; larger simulations would provide narrower confidence intervals

3. **Single Data Source**: ED visits capture only a subset of disease burden; community-based illnesses may be underrepresented

4. **Temporal Extrapolation**: Future outbreaks may differ in shape from the sigmoidal injection model used in sensitivity analysis

## Summary Recommendations

Based on the comparative evaluation:

- **GAM Adaptive** provides the best balance of accuracy and calibration for seasonal syndromes
- **Farrington Flexible** offers regulatory-standard methodology with established validation history
- **GLR-NB** is most sensitive to sudden shifts but may require tuning for local conditions
- **Poisson Baseline** confirms that sophisticated methods provide meaningful improvement over naive approaches

For operational deployment, a multi-model ensemble approach may provide the most robust surveillance, with concordance across methods increasing confidence in true alerts.

---

*Report generated: `r Sys.time()`*
