---
title: "Model Comparison for Syndromic Surveillance"
subtitle: "Systematic Evaluation of Detection Algorithms: `r params$area`"
author: "Andrea Pedot"
date: "`r Sys.Date()`"
params:
  area: "Lombardia"
output:
  html_document:
    theme: flatly
    toc: true
    toc-depth: 3
    code_folding: hide
    df-print: paged
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 14, fig.height = 12)
library(tidyverse)
library(mgcv)
library(surveillance)
library(kableExtra)
library(lubridate)
library(targets)
library(patchwork)
```

# Introduction

Syndromic surveillance is a public health strategy that monitors pre-diagnostic health indicators to detect disease outbreaks earlier than traditional laboratory-confirmed reporting. Rather than waiting for confirmed diagnoses, syndromic surveillance systems analyze patterns in Emergency Department (ED) visits, categorized by presenting symptoms or chief complaints. This approach enables early detection of unusual health patterns before laboratory confirmation, real-time situational awareness of population health trends, and proactive resource planning to anticipate healthcare system demands.

This analysis evaluates syndromic surveillance data from **`r params$area`**, covering ED visits from December 2022 through March 2026. The data encompasses ten syndrome categories:

- `tot`: Total ED visits (baseline denominator)
- `resp`: Respiratory conditions
- `polm`: Pulmonary conditions
- `ili`: Influenza-like illness
- `gastro`: Gastrointestinal conditions
- `alt_cosc`: Altered consciousness
- `conv`: Convulsions
- `dec_gravi`: Unexpected deaths
- `int_farm`: Drug intoxications
- `rash_cut`: Cutaneous rashes

## Current Surveillance System

The existing surveillance infrastructure in Lombardy employs a combination of methods:

- **Farrington Flexible Algorithm**: Applied to respiratory syndromes (`ili`, `polm`, `resp`) for detecting aberrations against historical baselines
- **EWMA (Exponentially Weighted Moving Average)**: Used alongside Farrington for respiratory syndrome monitoring, providing sensitivity to gradual shifts
- **Historical Min-Max Comparison**: A simple benchmark comparing current counts against the same epidemiological week in previous years, applied across all syndromes

## State of the Art

The European Centre for Disease Prevention and Control (ECDC) has developed the **EpiSignal Detection** tool as a reference implementation for automated outbreak detection. This framework, available through the `surveillance` R package, provides standardized algorithms including:

- **Farrington Flexible**: The Noufaily-adjusted quasi-Poisson regression method with reweighting for past outbreaks, trend estimation, and power transformation to handle overdispersion
- **GLR for Negative Binomial (GLRNB)**: A CUSUM-based Generalized Likelihood Ratio test that accumulates evidence for mean shifts in count data, designed for prospective surveillance with control over false alarm rates

These methods represent the current consensus for syndromic surveillance in European public health practice.

## Objective

This report systematically evaluates four statistical approaches for anomaly detection against Lombardy ED syndromic surveillance data: a naive Poisson baseline, an adaptive GAM with 14-day calibration, the Farrington Flexible algorithm, and the GLR Negative Binomial CUSUM. Models are compared across accuracy, calibration, historical alert patterns, and sensitivity to synthetic outbreak detection.

```{r load-data}
# Load all required data from targets
clean_data <- tar_read(clean_data)
all_data_list <- tar_read(all_data_list)
all_stats_list <- tar_read(all_stats_list)
all_baselines_list <- tar_read(all_baselines_list)

# Model results
val_results <- tar_read(val_results)
farrington_results <- tar_read(farrington_results)
glrnb_results <- tar_read(glrnb_results)
poisson_results <- tar_read(poisson_results)

# Sensitivity results
sens_results_gam <- tar_read(sens_results_gam)
sens_results_farrington <- tar_read(sens_results_farrington)

# Filter for current area
df_long <- all_data_list[[params$area]]

cat("Analysis Area:", params$area, "\n")
cat("Date Range:", as.character(min(df_long$Data_inizio_sett)), "to", 
    as.character(max(df_long$Data_inizio_sett)), "\n")
cat("Total Observations:", nrow(df_long))
```


# Setup

## Data Infrastructure

### Temporal Structure

The input data consists of rolling 7-day windows with daily overlap. Each row represents a week-long aggregation of ED visits, advancing by one day. This structure provides smooth temporal resolution while maintaining sufficient statistical power for rare syndromes.

```{r data-structure}
df_long |>
  group_by(sindrome) |>
  summarise(
    First_Date = min(Data_inizio_sett),
    Last_Date = max(Data_inizio_sett),
    N_Observations = n(),
    Mean_Cases = round(mean(casi, na.rm = TRUE), 1),
    SD_Cases = round(sd(casi, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  kable(caption = "Data Summary by Syndrome") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Train-Test Split Strategy

To prevent data leakage and ensure valid out-of-sample evaluation:

- **Training Period**: December 2022 – March 2025 (~27 months)
- **Testing Period**: March 2025 – March 2026 (12 months)

All model parameters are estimated exclusively from training data.

```{r split-definition}
validation_start <- max(df_long$Data_inizio_sett) - days(365)

train_df <- df_long |> filter(Data_inizio_sett <= validation_start)
test_df <- df_long |> filter(Data_inizio_sett > validation_start)

cat("Training observations:", nrow(train_df), "\n")
cat("Testing observations:", nrow(test_df), "\n")
cat("Validation start date:", as.character(validation_start))
```


## Exploratory Data Analysis

### Long-Term Trends

```{r eda-longterm}
#| fig-width: 14
#| fig-height: 12
ggplot(df_long, aes(x = Data_inizio_sett, y = casi, color = sindrome)) +
  geom_line(alpha = 0.6) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE, linewidth = 1) +
  facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Long-Term Syndrome Trends",
    subtitle = "Smoothed GAM trend overlaid on raw weekly counts",
    x = "Date", y = "Weekly Cases"
  )
```

### Seasonal Patterns

```{r eda-seasonal}
#| fig-width: 14
#| fig-height: 12
df_long |>
  mutate(month = month(Data_inizio_sett, label = TRUE)) |>
  group_by(sindrome, month) |>
  summarise(mean_cases = mean(casi, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month, y = mean_cases, group = sindrome, color = sindrome)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Average Seasonal Patterns by Month",
    x = "Month", y = "Mean Weekly Cases"
  )
```

### Overdispersion Analysis

Overdispersion occurs when variance exceeds the mean, violating Poisson assumptions.

```{r eda-overdispersion}
stats_current <- all_stats_list[[params$area]]
overdisp_stats <- stats_current$overdispersion

overdisp_stats |>
  mutate(Overdispersed = ifelse(overdisp_index > 1, "Yes", "No")) |>
  kable(digits = 2, caption = "Overdispersion Diagnostics by Syndrome") |>
  kable_styling(bootstrap_options = c("striped", "hover")) |>
  column_spec(5, bold = TRUE, 
              color = ifelse(overdisp_stats$overdisp_index > 1, "red", "green"))
```

### Autocorrelation Structure

The autocorrelation function (ACF) measures the correlation between observations at different time lags. Since our data consists of rolling 7-day windows advancing by 1 day, the lags represent **days**, not weeks. The prominent dip at lag 7 reflects the structural autocorrelation introduced by the rolling window: observations 7 days apart share no overlapping data points.

```{r eda-acf}
#| fig-width: 14
#| fig-height: 12
acf_data <- stats_current$acf

ggplot(acf_data, aes(x = lags, y = acf_vals, color = sindrome)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "red", alpha = 0.5) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~sindrome, ncol = 3, scales = "free_y") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Autocorrelation Function (ACF) of Daily Changes",
    subtitle = "Lags in days; dip at lag 7 reflects non-overlapping rolling windows",
    x = "Lag (days)", y = "ACF"
  )
```


## Model Specifications

```{r prepare-all-models}
# Standardize all model results upfront
gam_std <- standardize_gam_results(val_results, params$area)
farr_std <- standardize_farrington_results(farrington_results, params$area)
poisson_df <- poisson_results[[params$area]]
glrnb_df <- glrnb_results[[params$area]]

# Get dates for validation period
validation_start <- max(df_long$Data_inizio_sett) - days(365)
```

### Model 0: Poisson Baseline (Benchmark)

The simplest possible surveillance model uses the historical training mean as the expected value for all future observations:

$$\hat{\lambda} = \frac{1}{n_{\text{train}}} \sum_{t \in \text{train}} y_t$$

Thresholds are set using Poisson quantiles: $Q_{0.95}(\lambda)$ for yellow alerts and $Q_{0.99}(\lambda)$ for red alerts. This model ignores seasonality, trends, and overdispersion, serving as a lower-bound benchmark against which more sophisticated methods must demonstrate improvement.

```{r poisson-burst-plot}
#| fig-width: 14
#| fig-height: 12
if (!is.null(poisson_df) && nrow(poisson_df) > 0) {
  
  plot_data <- poisson_df |>
    mutate(
      Point_Status = case_when(
        Observed > Limit_99 ~ "Critical",
        Observed > Limit_95 ~ "Alert",
        TRUE ~ "Normal"
      )
    )
  
  ggplot(plot_data, aes(x = Date)) +
    geom_ribbon(aes(ymin = 0, ymax = Limit_95, fill = "Safe"), alpha = 0.3) +
    geom_ribbon(aes(ymin = Limit_95, ymax = Limit_99, fill = "Watch"), alpha = 0.3) +
    geom_ribbon(aes(ymin = Limit_99, ymax = Inf, fill = "Critical"), alpha = 0.2) +
    geom_line(aes(y = Observed), color = "black", linewidth = 0.6) +
    geom_point(data = filter(plot_data, Point_Status == "Alert"),
               aes(y = Observed), color = "orange", size = 1.8) +
    geom_point(data = filter(plot_data, Point_Status == "Critical"),
               aes(y = Observed), color = "darkred", size = 2.2) +
    facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = c("Safe" = "#2ca02c", "Watch" = "#ff7f0e", "Critical" = "#d62728")) +
    coord_cartesian(ylim = c(0, NA)) +
    theme_minimal() +
    labs(
      title = "Poisson Baseline: Retrospective Burst Detection",
      subtitle = "Static thresholds based on training mean",
      x = "Date", y = "Daily Cases", fill = "Risk Zone"
    ) +
    theme(legend.position = "bottom")
}
```


### Model 1: Adaptive GAM with Negative Binomial Distribution

A Generalized Additive Model (GAM) that decomposes ED visit counts into interpretable components using flexible spline functions. The model uses the Negative Binomial distribution to handle overdispersion.

**Flexible specification** (for syndromes with strong seasonality):
$$\log(\mu_t) = \beta_0 + f_1(\text{week\_num}_t; k=20) + f_2(\text{week\_of\_year}_t; k=12) + \boldsymbol{\gamma}^\top \text{dow}_t$$

**Stiff specification** (for sporadic syndromes without clear seasonal patterns):
$$\log(\mu_t) = \beta_0 + f_1(\text{week\_num}_t; k=3) + \boldsymbol{\gamma}^\top \text{dow}_t$$

Where:

- $f_1(\cdot; k)$ is a thin-plate regression spline with $k$ knots capturing long-term trends
- $f_2(\cdot; k)$ is a cyclic cubic spline for annual seasonality (ensuring smooth year-to-year transitions)
- $\text{dow}_t$ is a factor for day-of-week effects

**14-Day Adaptive Calibration**: To account for recent deviations from historical patterns (e.g., an unusually severe flu season), the model applies a multiplicative bias correction based on the trailing 14 days:

$$\text{bias}_t = \frac{\bar{y}_{(t-14):t}}{\bar{\hat{\mu}}_{(t-14):t}}$$

The adjusted prediction becomes $\tilde{\mu}_t = \text{bias}_t \cdot \hat{\mu}_t$, and thresholds are computed from $\text{NegBin}(\tilde{\mu}_t, \theta)$ where $\theta$ is the dispersion parameter estimated by the GAM.

```{r gam-burst-plot}
#| fig-width: 14
#| fig-height: 12
if (nrow(gam_std) > 0) {
  
  plot_data <- gam_std |>
    mutate(
      Point_Status = case_when(
        Observed > Limit_99 ~ "Critical",
        Observed > Limit_95 ~ "Alert",
        TRUE ~ "Normal"
      )
    )
  
  ggplot(plot_data, aes(x = Date)) +
    geom_ribbon(aes(ymin = 0, ymax = Limit_95, fill = "Safe"), alpha = 0.3) +
    geom_ribbon(aes(ymin = Limit_95, ymax = Limit_99, fill = "Watch"), alpha = 0.3) +
    geom_ribbon(aes(ymin = Limit_99, ymax = Inf, fill = "Critical"), alpha = 0.2) +
    geom_line(aes(y = Observed), color = "black", linewidth = 0.6) +
    geom_point(data = filter(plot_data, Point_Status == "Alert"),
               aes(y = Observed), color = "orange", size = 1.8) +
    geom_point(data = filter(plot_data, Point_Status == "Critical"),
               aes(y = Observed), color = "darkred", size = 2.2) +
    facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = c("Safe" = "#2ca02c", "Watch" = "#ff7f0e", "Critical" = "#d62728")) +
    coord_cartesian(ylim = c(0, NA)) +
    theme_minimal() +
    labs(
      title = "GAM Adaptive: Retrospective Burst Detection",
      subtitle = "Thresholds adjusted via 14-day adaptive calibration",
      x = "Date", y = "Daily Cases", fill = "Risk Zone"
    ) +
    theme(legend.position = "bottom")
}
```


### Model 2: Farrington Flexible Algorithm

The Farrington Flexible algorithm (Noufaily et al., 2012) is the ECDC-recommended method for aberration detection in count time series. It fits a quasi-Poisson GLM to historical reference data, with several refinements:

**Key parameters (deviations from defaults)**:

- `b = 2`: Uses 2 years of historical data as reference (default is 3)
- `w = 21`: Reference window of 21 days (3 weeks) around the current time point
- `noPeriods = 10`: Number of periods for seasonal adjustment
- `pastWeeksNotIncluded = 182`: Excludes the most recent 26 weeks (182 days) from reference to avoid contamination by ongoing outbreaks
- `reweight = TRUE, weightsThreshold = 2.58`: Downweights historical observations exceeding 2.58 standard deviations (past outbreaks)
- `powertrans = "2/3"`: Power transformation $y^{2/3}$ to stabilize variance
- `thresholdMethod = "nbPlugin"`: Uses negative binomial plug-in for threshold calculation

The algorithm produces time-varying thresholds that adapt to seasonal patterns and trend.

```{r farrington-burst-plot}
#| fig-width: 14
#| fig-height: 12
if (nrow(farr_std) > 0 && sum(!is.na(farr_std$Limit_99)) > 10) {
  
  plot_data <- farr_std |>
    filter(!is.na(Limit_95), !is.na(Limit_99)) |>
    mutate(
      Point_Status = case_when(
        Observed > Limit_99 ~ "Critical",
        Observed > Limit_95 ~ "Alert",
        TRUE ~ "Normal"
      )
    )
  
  ggplot(plot_data, aes(x = Date)) +
    geom_ribbon(aes(ymin = 0, ymax = Limit_95, fill = "Safe"), alpha = 0.3) +
    geom_ribbon(aes(ymin = Limit_95, ymax = Limit_99, fill = "Watch"), alpha = 0.3) +
    geom_ribbon(aes(ymin = Limit_99, ymax = Inf, fill = "Critical"), alpha = 0.2) +
    geom_line(aes(y = Observed), color = "black", linewidth = 0.6) +
    geom_point(data = filter(plot_data, Point_Status == "Alert"),
               aes(y = Observed), color = "orange", size = 1.8) +
    geom_point(data = filter(plot_data, Point_Status == "Critical"),
               aes(y = Observed), color = "darkred", size = 2.2) +
    facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = c("Safe" = "#2ca02c", "Watch" = "#ff7f0e", "Critical" = "#d62728")) +
    coord_cartesian(ylim = c(0, NA)) +
    theme_minimal() +
    labs(
      title = "Farrington Flexible: Retrospective Burst Detection",
      subtitle = "ECDC-standard algorithm with historical reweighting",
      x = "Date", y = "Daily Cases", fill = "Risk Zone"
    ) +
    theme(legend.position = "bottom")
}
```


### Model 3: GLR for Negative Binomial (CUSUM-based)

The Generalized Likelihood Ratio (GLR) test is a CUSUM-type method designed for prospective surveillance. Unlike Farrington which compares each observation independently to a baseline, GLR accumulates evidence over time, making it particularly sensitive to sustained increases.

The test statistic at time $t$ is:
$$\Lambda_t = \max\left(0, \Lambda_{t-1} + \log\frac{f(y_t; \mu_1, \alpha)}{f(y_t; \mu_0, \alpha)}\right)$$

Where:

- $f(y; \mu, \alpha)$ is the Negative Binomial probability mass function
- $\mu_0$ is the in-control mean (estimated from training data)
- $\mu_1 = e^\theta \cdot \mu_0$ is the out-of-control mean we want to detect
- $\alpha$ is the dispersion parameter (estimated via method of moments)
- $\theta = \log(1.5)$ means we're tuned to detect a 50% increase in the mean

An alarm triggers when $\Lambda_t > h$, where $h$ is chosen to control the Average Run Length (ARL) — the expected time until false alarm under in-control conditions. We use `c.ARL = 5` as the threshold.

**ARL (Average Run Length)**: The expected number of observations until an alarm is raised. Under in-control conditions, higher ARL means fewer false alarms; under out-of-control conditions, lower ARL means faster detection.

```{r glrnb-burst-plot}
#| fig-width: 14
#| fig-height: 12
if (!is.null(glrnb_df) && nrow(glrnb_df) > 0) {
  
  plot_data <- glrnb_df |>
    mutate(
      Point_Status = case_when(
        Alarm == 1 ~ "Critical",
        TRUE ~ "Normal"
      )
    )
  
  ggplot(plot_data, aes(x = Date)) +
    geom_line(aes(y = Observed), color = "black", linewidth = 0.6) +
    geom_line(aes(y = Expected), color = "blue", linetype = "dashed", alpha = 0.5) +
    geom_point(data = filter(plot_data, Point_Status == "Critical"),
               aes(y = Observed), color = "darkred", size = 2.2) +
    facet_wrap(~sindrome, scales = "free_y", ncol = 3) +
    coord_cartesian(ylim = c(0, NA)) +
    theme_minimal() +
    labs(
      title = "GLR-NB: Retrospective Burst Detection",
      subtitle = "CUSUM-based detection; red points = alarms; dashed blue = in-control mean",
      x = "Date", y = "Daily Cases"
    ) +
    theme(legend.position = "bottom")
}
```


# Results

## Residual Diagnostics

```{r prepare-diagnostics}
# Prepare data for all models
gam_diag <- gam_std |> 
  mutate(Residual = Observed - Expected) |>
  filter(!is.na(Std_Residual), !is.na(Expected))

farr_diag <- farr_std |> 
  filter(!is.na(Std_Residual), !is.na(Expected))

poisson_diag <- if (!is.null(poisson_df)) {
  poisson_df |> filter(!is.na(Residual), !is.na(Expected))
} else tibble()

glrnb_diag <- if (!is.null(glrnb_df)) {
  glrnb_df |> 
    mutate(Residual = Observed - Expected,
           Std_Residual = Residual / sqrt(Expected + 0.001)) |>
    filter(!is.na(Residual), !is.na(Expected))
} else tibble()
```

### Q-Q Plots {.tabset}

::: {.panel-tabset}

#### Poisson Baseline

```{r qq-poisson}
#| fig-width: 12
#| fig-height: 12
if (nrow(poisson_diag) > 0) {
  poisson_diag |>
    group_by(sindrome) |>
    mutate(
      observed = sort(Residual),
      theoretical = qnorm(ppoints(n()))
    ) |>
    ungroup() |>
    ggplot(aes(x = theoretical, y = observed)) +
    geom_point(alpha = 0.4, color = "#2ca02c", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Q-Q Plot: Poisson Baseline Residuals",
      subtitle = "Deviation from diagonal indicates departure from normality",
      x = "Theoretical Quantiles", y = "Sample Quantiles"
    )
}
```

#### GAM Adaptive

```{r qq-gam}
#| fig-width: 12
#| fig-height: 12
if (nrow(gam_diag) > 0) {
  gam_diag |>
    group_by(sindrome) |>
    mutate(
      observed = sort(Std_Residual),
      theoretical = qnorm(ppoints(n()))
    ) |>
    ungroup() |>
    ggplot(aes(x = theoretical, y = observed)) +
    geom_point(alpha = 0.4, color = "#1f77b4", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Q-Q Plot: GAM Adaptive Standardized Residuals",
      subtitle = "Deviation from diagonal indicates departure from normality",
      x = "Theoretical Quantiles", y = "Sample Quantiles"
    )
}
```

#### Farrington

```{r qq-farrington}
#| fig-width: 12
#| fig-height: 12
if (nrow(farr_diag) > 0) {
  farr_diag |>
    group_by(sindrome) |>
    mutate(
      observed = sort(Std_Residual),
      theoretical = qnorm(ppoints(n()))
    ) |>
    ungroup() |>
    ggplot(aes(x = theoretical, y = observed)) +
    geom_point(alpha = 0.4, color = "#d62728", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Q-Q Plot: Farrington Standardized Residuals",
      subtitle = "Deviation from diagonal indicates departure from normality",
      x = "Theoretical Quantiles", y = "Sample Quantiles"
    )
}
```

#### GLR-NB

```{r qq-glrnb}
#| fig-width: 12
#| fig-height: 12
if (nrow(glrnb_diag) > 0) {
  glrnb_diag |>
    group_by(sindrome) |>
    mutate(
      observed = sort(Residual),
      theoretical = qnorm(ppoints(n()))
    ) |>
    ungroup() |>
    ggplot(aes(x = theoretical, y = observed)) +
    geom_point(alpha = 0.4, color = "#9467bd", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Q-Q Plot: GLR-NB Residuals",
      subtitle = "Deviation from diagonal indicates departure from normality",
      x = "Theoretical Quantiles", y = "Sample Quantiles"
    )
}
```

:::

### Calibration Plots {.tabset}

::: {.panel-tabset}

#### Poisson Baseline

```{r calib-poisson}
#| fig-width: 12
#| fig-height: 12
if (nrow(poisson_diag) > 0) {
  ggplot(poisson_diag, aes(x = Expected, y = Observed)) +
    geom_point(alpha = 0.3, color = "#2ca02c", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    geom_smooth(method = "gam", color = "darkred", se = FALSE, linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Calibration Plot: Poisson Baseline",
      subtitle = "Perfect calibration follows the diagonal; red line shows actual trend",
      x = "Expected Cases", y = "Observed Cases"
    )
}
```

#### GAM Adaptive

```{r calib-gam}
#| fig-width: 12
#| fig-height: 12
if (nrow(gam_diag) > 0) {
  ggplot(gam_diag, aes(x = Expected, y = Observed)) +
    geom_point(alpha = 0.3, color = "#1f77b4", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    geom_smooth(method = "gam", color = "darkred", se = FALSE, linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Calibration Plot: GAM Adaptive",
      subtitle = "Perfect calibration follows the diagonal; red line shows actual trend",
      x = "Expected Cases", y = "Observed Cases"
    )
}
```

#### Farrington

```{r calib-farrington}
#| fig-width: 12
#| fig-height: 12
if (nrow(farr_diag) > 0) {
  ggplot(farr_diag, aes(x = Expected, y = Observed)) +
    geom_point(alpha = 0.3, color = "#d62728", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    geom_smooth(method = "gam", color = "darkblue", se = FALSE, linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Calibration Plot: Farrington",
      subtitle = "Perfect calibration follows the diagonal; blue line shows actual trend",
      x = "Expected Cases", y = "Observed Cases"
    )
}
```

#### GLR-NB

```{r calib-glrnb}
#| fig-width: 12
#| fig-height: 12
if (nrow(glrnb_diag) > 0) {
  ggplot(glrnb_diag, aes(x = Expected, y = Observed)) +
    geom_point(alpha = 0.3, color = "#9467bd", size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
    geom_smooth(method = "gam", color = "darkred", se = FALSE, linewidth = 1) +
    facet_wrap(~sindrome, scales = "free", ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Calibration Plot: GLR-NB",
      subtitle = "Perfect calibration follows the diagonal; red line shows actual trend",
      x = "Expected Cases", y = "Observed Cases"
    )
}
```

:::

### Residual Autocorrelation {.tabset}

::: {.panel-tabset}

#### Poisson Baseline

```{r acf-poisson}
#| fig-width: 14
#| fig-height: 12
if (nrow(poisson_diag) > 0) {
  acf_poisson <- poisson_diag |>
    group_by(sindrome) |>
    arrange(Date) |>
    summarise(
      acf_vals = list({
        r <- Residual
        if (length(r) > 15) as.numeric(acf(r, lag.max = 10, plot = FALSE)$acf[-1])
        else rep(NA_real_, 10)
      }),
      lags = list(1:10),
      .groups = "drop"
    ) |>
    unnest(c(acf_vals, lags))
  
  ggplot(acf_poisson, aes(x = lags, y = acf_vals)) +
    geom_hline(yintercept = 0, color = "gray50") +
    geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "red", alpha = 0.5) +
    geom_line(color = "#2ca02c", linewidth = 1.2) +
    geom_point(color = "#2ca02c", size = 3) +
    facet_wrap(~sindrome, ncol = 3, scales = "free_y") +
    theme_minimal(base_size = 14) +
    labs(
      title = "Residual Autocorrelation: Poisson Baseline",
      subtitle = "Values near zero indicate good temporal fit",
      x = "Lag (days)", y = "ACF"
    )
}
```

#### GAM Adaptive

```{r acf-gam}
#| fig-width: 14
#| fig-height: 12
if (nrow(gam_diag) > 0) {
  acf_gam <- gam_diag |>
    group_by(sindrome) |>
    arrange(Date) |>
    summarise(
      acf_vals = list({
        r <- Residual
        if (length(r) > 15) as.numeric(acf(r, lag.max = 10, plot = FALSE)$acf[-1])
        else rep(NA_real_, 10)
      }),
      lags = list(1:10),
      .groups = "drop"
    ) |>
    unnest(c(acf_vals, lags))
  
  ggplot(acf_gam, aes(x = lags, y = acf_vals)) +
    geom_hline(yintercept = 0, color = "gray50") +
    geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "red", alpha = 0.5) +
    geom_line(color = "#1f77b4", linewidth = 1.2) +
    geom_point(color = "#1f77b4", size = 3) +
    facet_wrap(~sindrome, ncol = 3, scales = "free_y") +
    theme_minimal(base_size = 14) +
    labs(
      title = "Residual Autocorrelation: GAM Adaptive",
      subtitle = "Values near zero indicate good temporal fit",
      x = "Lag (days)", y = "ACF"
    )
}
```

#### Farrington

```{r acf-farrington}
#| fig-width: 14
#| fig-height: 12
if (nrow(farr_diag) > 0) {
  acf_farr <- farr_diag |>
    group_by(sindrome) |>
    arrange(Date) |>
    summarise(
      acf_vals = list({
        r <- Residual
        if (length(r) > 15) as.numeric(acf(r, lag.max = 10, plot = FALSE)$acf[-1])
        else rep(NA_real_, 10)
      }),
      lags = list(1:10),
      .groups = "drop"
    ) |>
    unnest(c(acf_vals, lags))
  
  ggplot(acf_farr, aes(x = lags, y = acf_vals)) +
    geom_hline(yintercept = 0, color = "gray50") +
    geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "red", alpha = 0.5) +
    geom_line(color = "#d62728", linewidth = 1.2) +
    geom_point(color = "#d62728", size = 3) +
    facet_wrap(~sindrome, ncol = 3, scales = "free_y") +
    theme_minimal(base_size = 14) +
    labs(
      title = "Residual Autocorrelation: Farrington",
      subtitle = "Values near zero indicate good temporal fit",
      x = "Lag (days)", y = "ACF"
    )
}
```

#### GLR-NB

```{r acf-glrnb}
#| fig-width: 14
#| fig-height: 12
if (nrow(glrnb_diag) > 0) {
  acf_glrnb <- glrnb_diag |>
    group_by(sindrome) |>
    arrange(Date) |>
    summarise(
      acf_vals = list({
        r <- Residual
        if (length(r) > 15) as.numeric(acf(r, lag.max = 10, plot = FALSE)$acf[-1])
        else rep(NA_real_, 10)
      }),
      lags = list(1:10),
      .groups = "drop"
    ) |>
    unnest(c(acf_vals, lags))
  
  ggplot(acf_glrnb, aes(x = lags, y = acf_vals)) +
    geom_hline(yintercept = 0, color = "gray50") +
    geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "red", alpha = 0.5) +
    geom_line(color = "#9467bd", linewidth = 1.2) +
    geom_point(color = "#9467bd", size = 3) +
    facet_wrap(~sindrome, ncol = 3, scales = "free_y") +
    theme_minimal(base_size = 14) +
    labs(
      title = "Residual Autocorrelation: GLR-NB",
      subtitle = "Values near zero indicate good temporal fit",
      x = "Lag (days)", y = "ACF"
    )
}
```

:::

### Residual Overdispersion

```{r overdispersion-all-models}
make_overdisp_data <- function(df, model_name) {
  if (is.null(df) || nrow(df) == 0) return(tibble())
  
  if (!"Residual" %in% names(df)) {
    if ("Observed" %in% names(df) && "Expected" %in% names(df)) {
      df <- df |> mutate(Residual = Observed - Expected)
    } else return(tibble())
  }
  
  df |>
    filter(!is.na(Residual)) |>
    group_by(sindrome) |>
    summarise(
      n_obs = n(),
      mean_residual = mean(Residual, na.rm = TRUE),
      var_residual = var(Residual, na.rm = TRUE),
      overdisp_index = var_residual / (abs(mean_residual) + 0.001),
      .groups = "drop"
    ) |>
    mutate(Model = model_name)
}

overdisp_gam <- make_overdisp_data(gam_diag, "GAM Adaptive")
overdisp_farr <- make_overdisp_data(farr_diag, "Farrington")
overdisp_poisson <- make_overdisp_data(poisson_diag, "Poisson Baseline")
overdisp_glrnb <- make_overdisp_data(glrnb_diag, "GLR-NB")

overdisp_all <- bind_rows(overdisp_gam, overdisp_farr, overdisp_poisson, overdisp_glrnb)

if (nrow(overdisp_all) > 0) {
  overdisp_all |>
    select(sindrome, Model, var_residual, overdisp_index) |>
    pivot_wider(names_from = Model, values_from = c(var_residual, overdisp_index)) |>
    kable(digits = 2, caption = "Residual Variance and Overdispersion by Model") |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) |>
    scroll_box(width = "100%")
}
```


## Performance Metrics

```{r metrics-calculation}
metrics_gam <- calculate_model_metrics(gam_std, "GAM Adaptive")

metrics_farr <- if (nrow(farr_std) > 0) {
  calculate_model_metrics(farr_std, "Farrington")
} else tibble()

metrics_poisson <- if (!is.null(poisson_df) && nrow(poisson_df) > 0) {
  calculate_model_metrics(poisson_df, "Poisson Baseline")
} else tibble()

metrics_glrnb <- if (!is.null(glrnb_df) && nrow(glrnb_df) > 0) {
  glrnb_df |>
    group_by(sindrome) |>
    summarise(
      Model = "GLR-NB",
      N_Obs = n(),
      RMSE = sqrt(mean((Observed - Expected)^2, na.rm = TRUE)),
      MAE = mean(abs(Observed - Expected), na.rm = TRUE),
      MAPE = mean(abs(Observed - Expected) / (Observed + 1), na.rm = TRUE) * 100,
      Mean_Bias = mean(Observed - Expected, na.rm = TRUE),
      R2 = 1 - sum((Observed - Expected)^2, na.rm = TRUE) / 
           sum((Observed - mean(Observed, na.rm = TRUE))^2, na.rm = TRUE),
      Coverage_95 = NA_real_,
      Coverage_99 = NA_real_,
      Alert_Rate_95 = NA_real_,
      Alert_Rate_99 = mean(Alarm == 1, na.rm = TRUE) * 100,
      .groups = "drop"
    )
} else tibble()

all_metrics <- bind_rows(metrics_gam, metrics_farr, metrics_poisson, metrics_glrnb) |>
  filter(!is.na(RMSE))
```

### Summary Statistics

```{r metrics-summary}
all_metrics |>
  group_by(Model) |>
  summarise(
    Avg_RMSE = mean(RMSE, na.rm = TRUE),
    Avg_MAE = mean(MAE, na.rm = TRUE),
    Avg_R2 = mean(R2, na.rm = TRUE),
    Avg_Coverage_95 = mean(Coverage_95, na.rm = TRUE),
    Avg_Alert_Rate_99 = mean(Alert_Rate_99, na.rm = TRUE),
    .groups = "drop"
  ) |>
  kable(digits = 2, caption = "Model Performance Summary (Averaged Across Syndromes)") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### RMSE Comparison by Syndrome

```{r metrics-plot}
#| fig-width: 14
#| fig-height: 12
all_metrics |>
  ggplot(aes(x = RMSE, y = reorder(Model, -RMSE), fill = Model)) +
  geom_col() +
  facet_wrap(~sindrome, scales = "free_x", ncol = 3) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 14) +
  labs(
    title = "RMSE Comparison by Syndrome",
    subtitle = "Lower is better; x-axis varies by syndrome to highlight differences",
    x = "Root Mean Square Error", y = "Model"
  ) +
  theme(legend.position = "none")
```


## Historical Alert Reconstruction

```{r alert-tables}
gam_alerts <- gam_std |>
  filter(Status == "RED") |>
  select(Date, Syndrome = sindrome, Observed, Expected) |>
  mutate(Model = "GAM Adaptive")

farr_alerts <- farr_std |>
  filter(grepl("RED", Status)) |>
  select(Date, Syndrome = sindrome, Observed, Expected) |>
  mutate(Model = "Farrington")

poisson_alerts <- if (!is.null(poisson_df) && nrow(poisson_df) > 0) {
  poisson_df |>
    filter(Status == "RED") |>
    select(Date, Syndrome = sindrome, Observed, Expected) |>
    mutate(Model = "Poisson Baseline")
} else tibble()

glrnb_alerts <- if (!is.null(glrnb_df) && nrow(glrnb_df) > 0) {
  glrnb_df |>
    filter(Alarm == 1) |>
    select(Date, Syndrome = sindrome, Observed, Expected) |>
    mutate(Model = "GLR-NB")
} else tibble()

all_alerts <- bind_rows(gam_alerts, farr_alerts, poisson_alerts, glrnb_alerts)
```

### Alert Summary by Model

```{r alert-summary}
all_alerts |>
  group_by(Model, Syndrome) |>
  summarise(N_Alerts = n(), .groups = "drop") |>
  pivot_wider(names_from = Model, values_from = N_Alerts, values_fill = 0) |>
  kable(caption = "Number of Critical Alerts by Model and Syndrome") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Alert Timeline

```{r alert-timeline}
#| fig-width: 14
#| fig-height: 14
if (nrow(all_alerts) > 0) {
  all_alerts |>
    ggplot(aes(x = Date, y = Model, color = Model)) +
    geom_point(alpha = 0.7, size = 3) +
    facet_wrap(~Syndrome, ncol = 1, strip.position = "right") +
    scale_x_date(
      date_labels = "%b %Y", 
      date_breaks = "2 months",
      expand = c(0.02, 0.02)
    ) +
    scale_color_brewer(palette = "Set1") +
    theme_bw(base_size = 14) +
    theme(
      legend.position = "none",
      strip.background = element_rect(fill = "gray90", color = "gray50"),
      strip.text = element_text(face = "bold", size = 12),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "gray50", linewidth = 0.8),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      panel.spacing = unit(0.3, "lines")
    ) +
    labs(
      title = "Alert Timeline by Model",
      subtitle = "Each point represents a critical (99%) alert",
      x = "Date", y = NULL
    )
}
```

### Alert Concordance

```{r alert-concordance}
if (nrow(all_alerts) > 0) {
  concordance <- all_alerts |>
    group_by(Date, Syndrome) |>
    summarise(
      Models_Alerting = paste(sort(unique(Model)), collapse = ", "),
      N_Models = n_distinct(Model),
      .groups = "drop"
    ) |>
    filter(N_Models >= 2) |>
    arrange(desc(N_Models), desc(Date))
  
  if (nrow(concordance) > 0) {
    concordance |>
      head(50) |>
      kable(caption = "Alert Concordance: Dates with Multiple Models Alerting") |>
      kable_styling(bootstrap_options = c("striped", "hover")) |>
      scroll_box(height = "400px")
  } else {
    cat("No dates where multiple models simultaneously triggered alerts.")
  }
}
```


## Sensitivity Analysis: Probability of Detection

### GAM Adaptive

```{r sens-gam}
#| fig-width: 14
#| fig-height: 12
sens_gam_df <- sens_results_gam[[params$area]]

if (!is.null(sens_gam_df) && nrow(sens_gam_df) > 0) {
  ggplot(sens_gam_df, aes(x = Total_Cases, y = POD, color = factor(Duration_Days))) +
    geom_hline(yintercept = 0.8, linetype = "dotted", color = "gray40") +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5, alpha = 0.7) +
    scale_x_log10(labels = scales::comma) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
    scale_color_brewer(palette = "Dark2", name = "Outbreak\nDuration (days)") +
    facet_wrap(~sindrome, ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "GAM Adaptive: Sensitivity Analysis",
      subtitle = "Probability of Detection vs. Outbreak Magnitude",
      x = "Total Outbreak Cases (log scale)", y = "Probability of Detection"
    )
} else {
  cat("GAM sensitivity results not available.")
}
```

### Farrington Flexible

```{r sens-farrington}
#| fig-width: 14
#| fig-height: 12
sens_farr_df <- sens_results_farrington[[params$area]]

if (!is.null(sens_farr_df) && nrow(sens_farr_df) > 0) {
  ggplot(sens_farr_df, aes(x = Total_Cases, y = POD, color = factor(Duration_Days))) +
    geom_hline(yintercept = 0.8, linetype = "dotted", color = "gray40") +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5, alpha = 0.7) +
    scale_x_log10(labels = scales::comma) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
    scale_color_brewer(palette = "Dark2", name = "Outbreak\nDuration (days)") +
    facet_wrap(~sindrome, ncol = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = "Farrington Flexible: Sensitivity Analysis",
      subtitle = "Probability of Detection vs. Outbreak Magnitude",
      x = "Total Outbreak Cases (log scale)", y = "Probability of Detection"
    )
} else {
  cat("Farrington sensitivity results not available.")
}
```

### Comparative POD Summary

```{r sens-comparison}
get_pod80_threshold <- function(df, model_name) {
  if (is.null(df) || nrow(df) == 0) return(tibble())
  
  df |>
    group_by(sindrome, Duration_Days) |>
    arrange(Total_Cases) |>
    filter(POD >= 0.8) |>
    slice_head(n = 1) |>
    ungroup() |>
    mutate(Model = model_name) |>
    select(sindrome, Duration_Days, Threshold_80_POD = Total_Cases, Model)
}

pod80_gam <- get_pod80_threshold(sens_gam_df, "GAM Adaptive")
pod80_farr <- get_pod80_threshold(sens_farr_df, "Farrington")

pod80_combined <- bind_rows(pod80_gam, pod80_farr)

if (nrow(pod80_combined) > 0) {
  pod80_combined |>
    pivot_wider(names_from = Model, values_from = Threshold_80_POD) |>
    kable(caption = "Minimum Outbreak Size for 80% Detection Probability") |>
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

# Conclusion

This analysis implements strict temporal separation to prevent data leakage. All model parameters are estimated using data up to March 2025, with the subsequent 12 months serving exclusively as the test set. For validation metrics, models are re-trained at each time point using only past data, simulating real-world deployment conditions.

Uncertainty is quantified through Monte Carlo simulation, with sensitivity analysis using 50 random trials per magnitude level across multiple outbreak durations. Dual thresholds at 95% (Yellow) and 99% (Red) confidence levels provide graduated alert severity for operational decision-making.

Regarding generalizability, this evaluation applies state-of-the-art surveillance methods (the ECDC-recommended Farrington Flexible and GLR-NB algorithms)to real-world Emergency Department data from three distinct geographic areas within Lombardy. The syndrome-specific model selection (flexible vs. stiff GAM specifications) adapts to the epidemiological characteristics of each condition. However, results may not generalize to healthcare systems with different reporting structures or population demographics. Rare syndromes have limited statistical power for sensitivity estimation, and the rolling 7-day window structure introduces autocorrelation that must be considered when interpreting residual diagnostics.

Based on the comparative evaluation:

- **GAM Adaptive** provides a reasonable balance of accuracy and calibration for seasonal syndromes, with the 14-day bias correction enabling rapid adaptation to evolving baselines
- **Farrington Flexible** offers regulatory-standard methodology aligned with ECDC guidelines and established validation history across European surveillance systems. More years should be used, as it is meant to run with 5 years of data in its standard implementation.
- **GLR-NB** demonstrates high sensitivity to sustained shifts through its CUSUM-based accumulation, though operational deployment requires careful parameters tuning.
- **Poisson Baseline** confirms that at least some of the sophisticated methods evaluated here provide meaningful improvement over naive approaches.

For operational deployment, we recommend thorough parallel testing of multiple methods before full implementation across all ASSTs. Concordance across models increases confidence in true alerts, while discordance may indicate borderline signals requiring epidemiological review. Prospective validation over at least one complete seasonal cycle is advisable before relying on any single method for public health action.
---

*Report generated: `r Sys.time()`*
